Wait-free Queues with Polygarithmic Step Complexity

### 2022-03-23 article/psuedocode.tex
minimalize datastructure structure

\title{Wait-free Queues with Polygarithmic Step Complexity}
\date{\today}

@@

\begin{document}
\maketitle

%\begin{abstract}
%This is the paper's abstract \ldots
%\end{abstract}

\section{Previous Work}
\begin{table}[hbt]
  \begin{center}
  \begin{tabular}{l|c|c|c}
    & Type & Progress Property  & Complexity \\ \hline
    MichaelS96~\cite{DBLP:conf/podc/MichaelS96} & Singly Linked List & Lock-free & $\Omega(p)$ \\ \hline
    MoirNSS05~\cite{DBLP:conf/spaa/MoirNSS05} & Linked List with Elemination & & \\ \hline
    HoffmanSS07~\cite{DBLP:conf/opodis/HoffmanSS07} & Linked List with Baskets & & $\Omega(p)$ \\ \hline
    Ladan-MozesS08 ~\cite{DBLP:journals/dc/Ladan-MozesS08} & Doubly linked list & Lock-free & $\Omega(p)$ \\ \hline
    MilmanKLLP18~\cite{DBLP:conf/spaa/MilmanKLLP18} & Linked List with batching same type ops & Lock-free & \\ \hline
    GidenstamST10~\cite{DBLP:conf/opodis/GidenstamST10} & Linked list of arrays & Lock-free & $\Omega(p+len(array))$ \\ \hline
    KoganP11~\cite{DBLP:conf/ppopp/KoganP11} &  Linkedlist with head and tail pointers & Wait-free & $\textsc{O}(p\times\text{bakery-max})$
  \end{tabular}
  \caption{Linearizable ME-MD Queus}
  \end{center}
\end{table}

\section{ Universal Construction using Tournament Tree with Big CAS Objects }
\paragraph{}
Jayanti~\cite{DBLP:conf/podc/Jayanti98a} proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions. He also introduced~\cite{DBLP:conf/podc/ChandraJT98} a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses. Here, we first introduce a universal construction using $\textsc{O}(\log p)$ CAS operations~\cite{DBLP:conf/fsttcs/JayantiP05}. We use the universal construction as a stepping stone towards our queue algorithm, so we will not explain it in too much detail.
\paragraph{}
We introduce our universal construction in Algorithm~\ref{alg1} with a tournament tree with $p$ leaves and height $\log(p)$ is shared among $p$ processes. Nodes are CAS objects, and each leaf is assigned to one process exclusively. Each CAS object stores a sequence of operations. When process $P_i$ wishes to apply an operation $op$ on the implemented object, it appends $op$ to its assigned leaf and tries to propagate it up to the root. The sequence of operations stored in Node $n$'s CAS object shows the order of the operations propagated up to $n$.
The history of operations stored at the root is the linearization ordering. The operation $op$ is linearized when it is appended to the root.

\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}

Leaf $l_i$ stores the sequence of the operations invoked by $P_i$. 
The algorithm uses a subroutine \textsc{Refresh}($n$) that concatenates new operations from node $n$'s children (that have not already been propagated to $n$) to the sequence of operations stored in $n$ and tries to CAS the new sequence into $n$. In other words, \textsc{Refresh}($n$) tries to append $n$'s children's new operations to $n$'s sequence.
After a process adding a new operations to its leaf, it has to propagate new operations up to the root. \textsc{Propagate}($n$) tries to append $n$'s new operations to the root $n$ by recursively calling \textsc{Refresh}($n$). In each step if a \textsc{Refresh}($n$) fails, it means another CAS operation has succeeded; if so, it tries to \textsc{Refresh}($n$) again. If the second attempt fails too, another process has already appended the operations the current \textsc{Propagate} is trying to append.
Operations that were in $l_i$ before \textsc{Propagate}($l_i$.parent) was invoked are guaranteed to be added to the root by the time the \textsc{Propagate}($l_i$.parent) finishes.

\begin{figure}

\begin{center}

\begin{subfigure}[b]{.49\textwidth}

  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
\Tree [.{\begin{tabular}{|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c||c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c||c|c|}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
\end{tikzpicture}
}
  \caption{Operations after $||$ are new.}

\end{subfigure}
\hfill
\begin{subfigure}[b]{.49\textwidth}

  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ & $op_{l_4}$ & $op_{l_5}$ & $op_{r_3}$ & $op_{r_4}$  \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c|c|c||}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c|c||}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
\end{tikzpicture}
}
  \caption{New operations are added to the parent node.}

\end{subfigure}

\caption{\label{fig:ucexample} Propagate Step in Universal Construction}
\end{center}
\end{figure}

\begin{algorithm}
\caption{Universal Construction Idea}\label{alg1}
\begin{algorithmic}[1]
\begin{multicols}{2}

\Function{response}{Do}{\textsl{operation} op, \textsl{pid} i} 
\State \texttt{l\textsubscript{i}.\Call{append}{op}}
\State \Call{Propagate}{parent of l\textsubscript{i}}
\State \texttt{Run} \textsf{the sequence stored in root}
\State \texttt{\Return op}\textsf{'s response from line 4}
\EndFunction{Do}

\Function{void}{Propagate}{\textsl{node} n}
\If{\texttt{n==root}} \Return
\ElsIf{!\Call{Refresh}{n}}
\State \Call{Refresh}{$n$} \EndIf
\State \Call{Propagate}{parent of n}
\EndFunction{Propagate}

\columnbreak

\Function{boolean}{Refresh}{\textsl{node} n}
\State \texttt{old=} \Call{Read}{n}
\State \texttt{new=} \textsf{ops that n's children contain but \texttt{old} does not}
\State \texttt{new= old$\cdot$new}
\State \Return n.\Call{CAS}{old, new}
\EndFunction{Refresh}

$\textsc{O}(\log n)$ CAS operations are invoked to do a \textsc{Propagate}, but the CAS words store sequences of unbounded length.
The problem is that we are trying to store unbounded sequence of operations in each node $n$ (see Figure~\ref{fig::uc}). However, to compute the result of an operation, we only use the total ordering that is stored at the root. Although we use a similar construction for our queue implementation, we develop an implicit representation of the sequence of operations, so that we can use reasonable size CAS objects and still achieve polygarithmic step complexity.

\begin{figure}[h]
\begin{center}
\begin{tikzpicture}[level 1/.style={level distance=1.8cm,sibling distance=1cm}, level 2/.style={level distance=1.4cm,sibling distance=1cm}]

\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c}  \hline  $op_2^1$ & $op_2^2$ & $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_1^1$ & $op_4^2$ & $op_1^2$ \\ \hline\end{tabular}}  [.{\begin{tabular}{|c|c|c|c|c|c}  \hline $op_2^1$ & $op_2^2$ & $op_1^1$ & $op_1^2$ \\ \hline\end{tabular}}
      {\begin{tabular}{|c|c|c|c}  \hline $op_1^1$ & $op_1^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_2^1$ & $op_2^2$ \\ \hline\end{tabular}} ] [.{\begin{tabular}{|c|c|c|c|c|c}  \hline $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_4^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_3^1$ & $op_3^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_4^1$ & $op_4^2$ \\ \hline\end{tabular}} ] ]

\end{tikzpicture}
\caption{\label{fig::uc} Universal Construction: $op_j^i$ denotes the $i$th operation from process $j$. In each node, we store the ordering of all the operations propagated up to it.}
\end{center}
\end{figure}

\section{Block Tree}

%\subsection{Idea in a nutshell}
We apply two ideas from universal construction to create a new linearizable data structure agreeing on a sequence of elements among processes. First, there is a shared tournament tree among processes, in which each process appends its element to its leaf in the tree and then tries to propagate it up to the root by performing \textsc{Refresh()} operations at each node. Second, each operation is linearized when its element is appended to the root.

\subsection{Sequence of propagated operations}
\paragraph{}
The basic idea behind the universal construction is to create a linearization of operations invoked by processes. If we design a fast shared data structure in which processes can append elements to a sequence, we can use it to implement practical, fast implementaion of an object $O$ by using the sequence data structure to keep track of the sequence of operations  on $O$. In the following sections, we introduce our solution called Block~Tree.

%\begin{lemma}
%    Consider an instance of \textsc{Propagate(}\textnormal{n}\textsc{)}. When it terminates all operations that were in $n$ before \textsc{Propagate(}\textnormal{n}\textsc{)} will be in the root. \texttt{(not suitable place)}
%\end{lemma}

\subsection{Sequence of Sets of Concurrent Operations \label{subsec::block}}
\paragraph{}
 In the universal construction, we order new concurrent operations at each \textsc{Refresh}() and maintain that order in the path up to the root. However, we can instead keep track of sets of concurrent operations and create the total ordering of all operations at the root (see Figure~\ref{fig::set}).

 \begin{figure}[h]
\begin{center}
\begin{tikzpicture}

\Tree [.{$\big\{op_2^1,op_2^2,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_1^1, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_2^2\big\},\big\{op_1^1\big\},\big\{op_1^2\big\}...$ }
      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ... \\ \hline\end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ... \\ \hline\end{tabular}} ] ]

\end{tikzpicture}
\caption{\label{fig::set} In each internal node, we store the set of all the operations propagated together, and one can arbitrarily linearize the sets of concurrent operations among themselves. Since we linearize operations when they are added to the root, ordering the blocks in the root is important.}
\end{center}
\end{figure}

The definition of linearizability allows concurrent operations to be reordered arbitrarily. Thus, a group of concurrent operations can be appended to our root sequence as one block without specifying the order among the operations.

%\begin{lemma}
%Operations in the same block can be linearized in any order.
%\end{lemma}

\subsection{Using arrays instead of big CAS Objects}
\paragraph{}
 We used unbounded CAS objects storing sequences as big words in the universal construction. One can represent sequences as arrays to overcome this implementation problem. Each array element will store one of the blocks of concurrent operations described in section \ref{subsec::block}.

\subsection{Augmenting Tree to make \textsc{Refresh}() Step faster}
\paragraph{}
Copying operation sequences from children to their parent in a \textsc{Refresh}() takes time proportioned to the number of operations being copied. This is time-consuming, so we propose a way to augment the tree to calculate lines 15,16 in $\textsc{O}(\log p)$ steps which reads new operation and concats them with old operations. Instead of representing the set of operations by explicitly listing them in a node, we represent a set of operations implicitly by recording which of the children's sets were unioned to create the set. Having operation sequences stored at leaves, we can deduce a set of operations in a node using this implicit representation.~(see Figure~\ref{fig:block}.)

\begin{figure}

\begin{center}
\begin{tikzpicture}[level 1/.style={level distance=3.3cm,sibling distance=1cm},
	level 2/.style={level distance=2.5cm,sibling distance=0.5cm},
	level 3/.style={level distance=1.8cm,sibling distance=1.2cm}]

\Tree [.{\begin{tabular}{||c|c|c||}  \hline 0,11 & 12,6 & 8,36 \\ \hline\end{tabular}}
 [.{\begin{tabular}{||c||c||}  \hline 10,2 & 5,3 \\  \hline \multicolumn{1}{c||}{2} & \multicolumn{1}{c}{3} \end{tabular}}
 [.{\begin{tabular}{||c|c||c||}\hline 2,3 & 4,1 & 0,5 \\\hline\multicolumn{2}{c||}{1} & \multicolumn{1}{c}{2}\end{tabular}} $\vdots$ $\vdots$ ]
  [.{\begin{tabular}{||c||c||}  \hline 0,2 & 1,2 \\  \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c}{2} \end{tabular}} $\vdots$ $\vdots$ ] ]
          [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2} & \multicolumn{2}{c}{3}\end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2}& \multicolumn{2}{c}{3}\end{tabular}} $\vdots$ $\vdots$ ]
            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline \multicolumn{2}{c||}{1} & \multicolumn{1}{c||}{3} & \multicolumn{2}{c}{4}\end{tabular}} $\vdots$ $\vdots$ ] ] ]

\end{tikzpicture}
\end{center}
\caption{\label{fig:block} Showing concurrent operation sets with blocks. Each block consists of a pair(left, right) indicating the number of operations from the left and the right child, respectively. Block (12,6) in the root contains blocks (10,2) from the left child and (6,0) from the right child. Blocks between two lines $||$ are propagated together to the parent. For example, Blocks (2,3) and (4,1) from the leftmost leaf and (0,2) from its sibling are propagated together into the block (10,2) in their parent. The number underneath a group of blocks in a node indicates which block in the node's parent those blocks were propagated to.}
\end{figure}

Each block $b$ in node $n$ is the aggregation of blocks in the  children of $n$ that are newly read by the\textsc{Propagate}() step that created block $b$. For example, the third block in the root (8,36) is created by merging block (5,3) from the left child and (14,6) and (0,16) from the right child. Block (5,3) also points to elements from blocks (0,5) and (1,2). 
%Recursively we can deduce each block contents until the leaf nodes store arrays of individual element operations.

\begin{definition}\{Existence of an operation in a block\}  Operation op exists in block b if it has propagated up to block b.
\end{definition}

\begin{definition}\{Subblock\}
  The blocks that are aggregated into block $b$ in a \textsc{Propagate}() step are called subblocks of $b$. Block $b_1$ is a subblock of $b_2$ if and only if $b_1$ is a block in node $v$ and in $b_2$ is a block in the parent of $v$ and $b_1$'s elements exits in $b_2$'s elements.
\end{definition}
%{\TODO: describe how blocks work with a good example how exactly it is done}

\paragraph{}
We choose to linearize operations in a block from the left child before those from the right child as a convention. Operations within a block of the root can be ordered in any way that is convenient. In effect, this means that if there are concurrent new blocks in a \textsc{Refresh}() step from several processes we linearize them in the order of their process ids. So for example  operations aggregated in block (10,2) are in the order (2,3),(4,1),(0,2). All blocks from the left child with come before the right child and the order of blocks of each child is preserved among themselves.
%\begin{lemma}
%\label{lem:block_size}
%There cannot be more than one operation from each process in a \textsc{Refresh}() invocation.
%\end{lemma}
%\begin{proof}
%  If so, then that parent process has invoked two simultaneous operations. $ \blacksquare $
%\end{proof}

\paragraph{}
In a \textsc{Propagate}() invocation path from a leaf to root, there will be \textsc{Refresh}() steps with merges from $2, 4, 8, ..., p$ processes. So in a complete propagation, at most $2p$ blocks are merged into one block. \texttt{(maybe useful for analysis)}
%\begin{lemma}
%A block in the root cannot contain more than one operation from a process.
%\end{lemma}
%\begin{proof}
%  Assume a process in a block invokes two operations; one has to be finished another. If so, it has been added to root before. $\blacksquare$
%\end{proof}

\subsection{Using pointers and prefix sum to make \textsc{GetIndex}($i$) faster}
%\begin{definition}
%  \textsc{BSearch}($i$): Search $i$th element in a prefix sum array with length $n$ in \textsc{O}($\log n$) steps.
%\end{definition}
\paragraph{}
\textsc{GetIndex}($i$) returns the $i$th operation stored in the block tree sequence. We do that by finding the block $b_i$ containing $i$th element in the root, and then recursively finding the subblock of $b_i$ which contains $i$th element. To make this recursive search faster, instead of iterating over all elements in sequence of blocks we store prefix sum of number of elements in the blocks sequence and pointers to make BSearch faster.

Furthermore, in each block, we store the prefix sum of left and right elements. Moreover, for each block, we store two pointers to the last left and right subblock of it (see fig \ref{fig::pointer} and \ref{fig:prefix}).

\begin{figure}
\begin{center}
\begin{tikzpicture}[level 1/.style={level distance=2.3cm,sibling distance=1cm}]

\Tree [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline  0,0 & 3,8 & 9,8 & 23,14\end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline 0,0 & 0,3 & 4,5 & 9,6 \end{tabular}} ]
            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline 0,0 & 2,1 & 7,1 & 11,3 & 14,7 \end{tabular}} ] ]

\end{tikzpicture}
\end{center}
\caption{\label{fig:prefix} Using Prefix sums in blocks. When we want to find block b elements in its children, we can use binary search. The number below each block shows the count of elements in the previous blocks.}
\end{figure}

\begin{figure}[hbt]
\centering
  \includegraphics[width=5in]{pics/pointers}
  \caption{Block have pointers to the starting block of theirs for each child. \label{fig::pointer}}
\end{figure}

\paragraph{}
Starting from the root, \textsc{GetIndex}($i$) BSearches $i$ in the prefix sum array to find block containing $i$th operation, then continues recursively calling \textsc{GetElement}($b,i$) to find $i$th element of block $b$. From lemma $\ref{lem:block_size}$ we know a block size is at most $p$. So BSearch takes at most \textsc(O)$(\log p)$, since  with knowing pointers of a block and its previous block we can determine the base \texttt{(domain ?)} to search and its size is \textsc{O}$(p)$.

\subsection{Block Tree Algorithm}
\paragraph{}
Our Block Tree is a linearizable implementation of a data structure that stores a sequence of elements. It has two methods (see Algorithm~\ref{alg2}), \textsc{Append}($e$) which appends element $e$ to the sequence, and \textsc{Get}($i$) which returns the $i$th element in the sequence.

\paragraph{Design of a block tree}

Each process is assigned to a leaf in a shared tournament tree. Thus, for example, the leaf node for process $p_i$ contains an array of elements by $p_i$ in the order they were invoked.
Each internal node of the tree contains an array of blocks of elements.
Block $b$ in node $n$ is created in a \textsc{Propagate}() step and is merged block of new blocks at the time of \textsc{Propagate}() reading $n$'s children blocks. Each block consists of pointers left and right, to the last block merged into itself from left and right child in that order. Moreover, two numbers, left and right, indicate the count of elements in the blocks from the left and right child consecutively. Furthermore, prefix left, and right can be computed from the prefix sum of left and right values.
Elements of block $b$ can be determined recursively (\textsc{GetElements($b$)}).
The $i$th element in the sequence can be determined in \textsc{O}($\log^2 p$) steps by recursively finding $i$th element in block $b$ (\textsc{GetElement($i$)})
After element $e$ is propagated (appended to a block int the root), its index can be computed with \textsc{GetIndex}($op$).

In order to compute elements of a block faster we store prefix-sum blocks(block i has tuple(right-sum=$\#$right ops in previous block, left-sum=$\#$left ops in previous blocks)[See Figure \ref{fig:prefix}]. Here is the algorithm to get elements of a block.

\paragraph{Specification}
A block tree is a shared data structure that stores a sequence of elements.  It has two methods \texttt{Append(e)} and \texttt{Get(i)}. \texttt{Append(e)} adds \texttt{e} to the end of the sequence and returns the index of \texttt{e} in the sequence. \texttt{Get(i)} returns \texttt{i}th element stored in the sequence.

\paragraph{SubBlock} Block \texttt{s} is a subblock of \texttt{b} if \texttt{s} is between blocks \texttt{start..end} in \texttt{n} from Lines 41,42 of \texttt{CreateBlock()}.

\paragraph{Membership}
Element \texttt{e} is a member of block \texttt{b} in:
\begin{itemize}
  \item internal node \texttt{n}, if \texttt{e} is a memeber of \texttt{s} that \texttt{s} is a subblock of \texttt{b}.
  \item leaf node \texttt{n}, if \texttt{e} belongs to \texttt{n.dir.blocks[b\textsuperscript{$\prime$}.end\textsubscript{dir}+1..b.end\textsubscript{dir}]} for \texttt{dir} $\in$ \texttt{\{left, right\}} which b\textsuperscript{$\prime$} is the previous block of \texttt{b} in \texttt{n}.
\end{itemize}

\paragraph{Order of elements inside node}
Element \texttt{d} is before element \texttt{e} in node \texttt{n}, if:
\begin{itemize}
  \item The block containing d is before the block containing e.
  \item e and d are in the same block and d is in the left child and e is in the right child.
  \item d is before e in the same child's order.
\end{itemize}

\begin{figure}[hbt]
  \center\includegraphics[width=5.5in]{pics/tree}
  \caption{Order of elements in b: elements in leaves are ordered with numerical order in the drawing.}
\end{figure}

\paragraph{\texttt{CreateBlock()}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children. \texttt{b\textsuperscript{$\prime$}.end\textsubscript{left}} stores the index of the rightmost subblock of left child of \texttt{b}'s previous block. Other attributes are assigned values followed by definition.
\begin{figure}[hbt]
  \center\includegraphics[width=5.5in]{pics/createblock}
  \caption{\label{fig::createBlock}Snapshot of a CreateBlock()}
\end{figure}

%\textsf{Is it necessary to say \texttt{CreateBlock()} is doing what it's supposed to do by definition? Or \texttt{Append()} for example?}

\paragraph{Double Refresh}
Elements in \texttt{n}'s children's blocks before line \texttt{13} are guaranteed to be in \texttt{n}'s blocks after Line~\texttt{15}.
\begin{proof}
\texttt{CreateBlock()} reads blocks in the children that does not exist in the parent and aggregates them into one block. If a \texttt{Refresh} procedure returns true it means it has appended the block created by \texttt{CreateBlock()} into the parent node's sequence. So suppose two \texttt{Refresh}es fail. Since the first \texttt{Refresh} was not successful, it means another CAS operation by a \texttt{Refresh}, concurrent to the first\texttt{Refresh}, was successful before the second \texttt{Refresh}. So it means the second failed \texttt{Refresh} is concurrent with a successful \texttt{Refresh} that assuredly has read block before the mentioned line \texttt{13}. After all it means if any of the \texttt{Refresh} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
\end{proof}
\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/doublerefresh.png}
  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh()} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock()}.}
\end{figure}

\paragraph{Disjunction} Blocks in node \texttt{n}'s contain disjoint sets of elements.
\begin{proof}
Without loss of generality, assume blocks \texttt{b1, b2} contain common element \texttt{e} from the left child, and \texttt{b2} is after \texttt{b1} in \texttt{n}'s sequence of blocks. So block start of b2's \texttt{CreateBlock()} is after block end of \texttt{b1}'s end. Since \texttt{b2}'s start is the end of the block before itself, it cannot be before \texttt{b1}'s end.
\end{proof}

\paragraph{Total Order}
Sequence represented by the Block Tree is the sequence of the blocks stored in the root.

\paragraph{Linearization Points}
\texttt{Get(i)} is linearized when it terminates. \texttt{Append(e)} is linearized right after when a block containing \texttt{e} is appended to the root, if there are multiple elements appended together, they are linearized by the defined order in the root.

\paragraph{Subblocks Upperbound} Block b has at most $p$ subblocks.
\begin{proof}
If there are more than $p$ subblocks, then there is more than one block from process pl. \texttt{Append(e)} finishes after propagating and appending \texttt{e} to the root(line 9). So these blocks cannot be appended to root already, so pl has invoked two concurrent \texttt{Append()}s(line 1) without terminating the first one.
\end{proof}

\paragraph{Computing \texttt{Get(n, b, i)}}
To find the \texttt{i}th element in block \texttt{b} of node \texttt{n}, we search among subblocks of \texttt{b} that is bounded by $p$. Subblocks of a block are within the start and end block of the \texttt{CreateBlock()} procedure of it.

\paragraph{How \texttt{Refresh(n)} works.}
\begin{enumerate}
  \item Read n's counter and head
  \item Create block b
  \item CAS b into n
  \item If previous succeed:
  \begin{enumerate}
  \item Update sup of b's ending subblocks
  \item Increment children's counters
  \end{enumerate}
\end{enumerate}

%\paragraph{Time-Space Relation} There are at most $p^2$ blocks with same \texttt{time} attribute.
%\begin{proof}
%If a process dies at line 28, the children counter won't increase, and each process aggregates most block, and since there are at most p concurrent processes, then there will be at most $p^2$ blocks with same \texttt{time} attribute.
%\end{proof}
%
%\paragraph{Time-Propagate() Relation} In a \texttt{CreateBlock()} procedure, the blocks that are seen have at most 2 different \texttt{time} attribute values.
%
%\paragraph{Time-Super Relation} sup[i+1]-sup[i] is $\leq p$
%
%\paragraph{\texttt{b}'s Superblock position} SuperBlock of \texttt{b} is within range $\pm p$ of the \texttt{\texttt{sup[b.time]}}.
%\begin{proof}
%
%\end{proof}
\begin{figure}[hbt]
  \center\includegraphics[width=6in]{pics/super}
  \caption{\texttt{Sup} and \texttt{timer} in a node, numbers on blocks are their \texttt{time} values.}
\end{figure}

\paragraph{Computing superblock}
\begin{enumerate}
  \item Value read for \texttt{super[b.time]} in line 71 is not null.\begin{proof}
    \texttt{Index()} is invoked after finishing \texttt{Propagate()} in line 10. For each value \texttt{c\textsubscript{dir}} read in lines 23, \texttt{super} is set before incrementing in lines 26,27.
  \end{proof} 
  \item \texttt{super[]} preserves order from child to parent; if in a child block \texttt{b} is before \texttt{c} then \texttt{b.time} $\leq$ \texttt{c.time} and \texttt{super[b.time]} $\leq$ \texttt{super[c.time]}\begin{proof}
   Follows from the order of lines 37, 26, 27.
 \end{proof}
%  \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
%  \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
  \item \texttt{super[i+1]-super[i]}$\leq p$
\begin{proof}
 In a Refresh with successful CAS in line 24, \texttt{super} and \texttt{counter} are set for each child in lines 26,27. Assume the current value of the counter in node \texttt{n} is \texttt{i+1} and still \texttt{super[i+1]} is not set. If an instance of successful \texttt{Refresh(n)} finishes \texttt{super[i+1]} is set a new value and a block is added after \texttt{n.parent[sup[i]]}. There could be at most $p$ successful unfinished concurrent instances of \texttt{Refresh()} that have not reached line 27. So the distance between \texttt{super[i+1]} and \texttt{super[i]} is less than $p$.
 \end{proof}
  \item Superblock of \texttt{b} is within range $\pm 2p$ of the \texttt{super[b.time]}.
\begin{proof}
\texttt{super[i]} is the index of the superblock of a block containing block b. It is trivial to see that \texttt{n.super} and \texttt{n.b.counter} are increasing.  \texttt{super(b)} is the real superblock of b. \texttt{super(t]} is the index of the superblock of the last block with time \texttt{t}. If \texttt{b.time} is \texttt{t} we have:
$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
  \end{proof} 
\end{enumerate}

%The block tree contains a tournament tree in which leaves are assigned to processes, and each process appends elements to its leaf and tries to propagate it up to the root. The final sequence is stored at the root. \texttt{Append} after appending to leaf calls \texttt{Propagate} which iterates over the path to root. In each \texttt{Propagate(n)} invocation on node n it tries ti \texttt{Refresh} the sequence stored at node n with the new elements in n's children. After adding the element to the root \texttt{Index(op)} finds the order of the element in the root.

%\textsf{I finally thought maybe a sequence of elements is better for telling specification.}

\begin{algorithm}
\caption{Block Tree \label{alg2}}
\begin{algorithmic}[1]
\begin{multicols}{2}

\Statex \textbf{Structure}

\Statex $\blacktriangleright$ \texttt{\textsl{element} e}
\begin{itemize}
%\item \texttt{\textsl{string} name}
\item \texttt{\textsl{int} pid}
\item \texttt{\textsl{int} loc\textsf{: location in \texttt{l\textsubscript{pid}}.ops}}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{node} n}
\begin{itemize}
\item \texttt{\textsl{*node} left, right, parent}
\item \texttt{\textsl{block[]} blocks}\textsf{: blocks stored in \texttt{n}}
\item \texttt{\textsl{int} head= 1}\textsf{: index of first empty cell of \texttt{blocks}}
\item \texttt{\textsl{int} counter= 0}\textsf{}
\item \texttt{\textsl{int[]} super}\textsf{: \texttt{super[i]} stores the index of a superblock in parent that contains some block of this node whose \texttt{time} is \texttt{i}}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{leaf node} l\textsubscript{i} extends \textsl{node}}
\begin{itemize}
\item \texttt{\textsl{operation[]} ops}\textsf{: elements that are invoked to \texttt{Append()} to the blok tree by process i}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{block} b}
\begin{itemize}
  \item \texttt{\textsl{int} num\textsubscript{left}, sum\textsubscript{left}}

  \textsf{~~\#operations from the left subblocks of b, prefix sum of num\textsubscript{left}}

  \item \texttt{\textsl{int} num\textsubscript{right}, sum\textsubscript{right}}

  \textsf{~\#operations from the right subblocks of b, prefix sum of num\textsubscript{right}}

  \item \texttt{\textsl{int} sum}

  \textsf{\# operation in b}

  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}

  \textsf{~~index of \texttt{b}'s last subblock}
  \item \texttt{\textsl{int} time}

%\Statex \textbf{Examples}
%\Statex \hspace{0pt} \texttt{ l\textsubscript{i}[j]}\textsf{: jth operation of ith process}
%\Statex \hspace{0pt} \texttt{ n.left}\textsf{: left child of node n}
%\Statex \hspace{0pt} \texttt{ n.blocks[n.head]}\textsf{: leftmost empty block of node n}
%\Statex \hspace{0pt} \texttt{ n.blocks[i].num\textsubscript{left}}\textsf{: \#left operations of ith block of n}
%\Statex \hspace{4pt} \texttt{b.end\textsubscript{right}}\textsf{: index of ending right subblock of b}

\Function{int}{Append}{\textsl{operation} op, \textsl{int} pid} 
\State \texttt{op.loc= l\textsubscript{pid}.head}
\State \texttt{block b= \Call{new}{\textsl{block}}}
\State \texttt{b.time= op.loc}
\State \texttt{b.sum= 1}
\State \texttt{l\textsubscript{pid}.ops[op.loc]= op}
\State \texttt{l\textsubscript{pid}.blocks[op.loc]= b}
\State \texttt{op.head+= 1}
\State \Call{Propagate}{l\textsubscript{pid}.parent}
\State \Return \Call{Index}{l\textsubscript{pid}, op.loc, b} 
\EndFunction{Append}

\columnbreak

\Function{void}{Propagate}{\textsl{node} n}
\If{\texttt{not} \Call{Refresh}{n}}
\State \Call{Refresh}{n}
\EndIf
\If{\texttt{n {\keywordfont is} root}} \Return
\EndIf
\State \Call{Propagate}{n.parent}
\EndFunction{Propagate}

\Function{boolean}{Refresh}{\textsl{node} n}
\State \texttt{i= n.head}
\State \texttt{c= n.counter}
\State \texttt{new, c\textsubscript{left}, c\textsubscript{right}= \Call{CreateBlock}{n, i+1, c}}
\If{\Call{CAS}{n.blocks[i+1], null, new}}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
\State \texttt{\Call{CAS}{n.dir.super[c\textsubscript{dir}], null, i+1}}
\State \texttt{\Call{CAS}{n.dir.counter, c\textsubscript{dir}, c\textsubscript{dir}+1}}
\EndFor
\State \Return{\texttt{true}}
\Else 
\State \Return{ \texttt{false}}
\EndIf
\State \texttt{\Call{CAS}{n.head, i, i+1}}
\EndFunction{Index}

>>>>

@@

\begin{document}
